### 2019-07-09 Manuscripts/post-cogpsych/post-cogpsych.tex
Modelling for hybrid rep+gcm models finally done. Fixed comments as suggested by Joe, also improved writing in some areas (I think). Ready to send to ken+nolan.

@@

Despite both the original hierarchical Bayesian model and our representativeness
model sharing identical hierarchical structures, their distinct response
processes can yield very different exemplars. Specifically, the hierarchical
Bayesian model emphasizes the generation of exemplars that maintain
distributional commonalities across categories. In contrast, the
representativeness model focuses on the generation of exemplars that are
representative of the underlying distribution for a target category. Generally,
with the assumption of unimodal underlying distributions, this mechanism of
representativeness results in the generation of novel exemplars that are less
similar to exemplars from the known category. This occurs because the
representativeness of novel exemplars tend to be highest where the probability
density of the known category's underlying distribution is the lowest. We see
this illustrated in Figure \ref{fig:representative-examples}, where the
representativeness model displays a strong preference for generating exemplars
from target category 'B' that are further away from the contrast category 'A'.
Exceptions to this pattern can occur (i.e., the generation of novel exemplars
similar to known category exemplars) when the known category distribution is
especially large on all features -- we explore this in more detail in Section
\ref{section:diffcon} where we highlight an important condition demonstrating
how PACKER and the representativeness model can make qualitatively different
predictions.

@@

For the representativeness model, adding an unbounded feature changes the probability distribution for the underlying Bayesian model. Whereas it is still Gaussian in one feature, the other feature is toroidal, meaning that it wraps back to $0$ radians once it reaches $2\pi$ (we have conducted a change of variables where $y \rightarrow y \mod 2\pi $, where $\mod$ is the modulus operator). As we cross $2\pi$ and reach $0$ again, the remaining probability density also gets wrapped around, meaning that the probability density assigned by the Gaussian distribution on the toroidal dimension at $2\pi$ is actually assigned to $0$ (and $2\pi + \epsilon$ to $\epsilon$ so on). This is called the {\em wrapped} Gaussian distribution \citep{mardia1972}. Note that the often used distribution for circular dimensions, the Von Mises distribution is an approximation to the wrapped Gaussian distribution. See Appendix B for implementational details.

@@

We recruited 89 participants from Amazon Mechanical Turk who were randomly assigned to the bounded, square stimuli ($N=46$), or to the toroidal, circle stimuli ($N=43$). The square stimuli were exactly the same as in Experiments 1 and 2. The circle stimuli were made up of unfilled circles with a radial line extending from its center to its edge. These stimuli varied along two features -- their
diameter ranging from 3.0--5.8 cm, and their orientation ranging from
0--$2\pi$ radians. The orientation of each circle stimuli was fixed at a
randomly selected angle for all participants. Similar to Experiments 1 and 2,
participants first observed a training phase where four Alpha exemplars were
presented across three blocks followed by the generation phase. The training condition was the Corner category for all participants.

@@

The contrast models were fit by optimizing their fit to all experiments in this article (Experiments 1 through 3). Fit statistics and parameter values are presented in Table \ref{table:corner-model-fits}. Overall,
the representativeness model emerges as the best-fitting model in aggregate, although by a much smaller amount than Experiments 1 and 2. However, analyzing data in aggregate can be misleading.

@@

\begin{tabular}{ l l} \\ \textbf{PACKER} & \textbf{Representativeness} \\
  \hline $AIC = 12052$ & $AIC = 11991$ \\ 
  $L = -6023$          & $L = -5991$ \\ 
  $c = 0.60$           & $\kappa = 5.91$  \\ 
  $\theta_c = 2.78$    & $\nu = 1.00$ \\
  $\theta_t = 3.17$    & $\lambda = 1.45$ \\
                       &$\theta = 1.86$ \\
\end{tabular}
\end{table}

@@

We conducted simulations comparing the contrast models' account of our results
to the classical proposals for category generation: a ``copy-and-tweak'' model
(realized as a variant of PACKER with no sensitivity to category contrast), and
a hierarchical Bayesian model designed to explain the classic distributional
similarity effect. In all simulations, we found that the contrast models
captured a previously unexplained and unexplored aspect of human category
generation. In particular, by measuring PACKER's fit as a function of its
prioritization of within- and between-category similarity, we observed that
considering either constraint exclusively results in a relatively low-quality
account. Instead, PACKER's best results were obtained when both constraints are
considered, indicating that human learners do not generate novel concepts
exclusively on the basis of within-category similarity or between
class-contrast. This finding mirrors our behavioral results and demonstrates
that both constraints influence generation when explained with an exemplar
model. Ultimately, our category generation data were best explained by the
representativeness model. The exact nature of its superiority over the other
models is still not yet clear, although we find evidence that it only barely
surpasses PACKER when considering known categories with extreme exemplars.
As we explore in a later section, it is likely that the
representativeness mechanism itself (i.e., independent of the hierarchical
Bayesian foundation) captures a core aspect of both category generation and
category learning.

@@

How well does a representativeness model that uses exemplar similarity to
represent the category’s density over features explain our experimental results?
We find that this model substantially outperforms its classical counterpart, but
not to the extent of PACKER or the representativeness model. Specifically,
fitting this model to the category generation data yielded $L = -4815$ and
$AIC=9633$ for the entire set of data from Experiments 1 and 2, and $L = -3474$,
$AIC=6953$ for data excluding the first trials. When used to predict the error
rates of participants in the category learning task (Experiment 3), the
performance of the copy-and-tweak with representativeness model correlated with
participant error with $r=.56, p < .001$. These results suggest that the
predictive advantages to including the representativeness mechanism is not
limited to a hierarchical Bayesian model, but can also be found when applied to
an exemplar model. Although its performance is not as strong as either of the
fully-developed contrast models, there is a benefit to the representativeness
mechanism that is independent from any interaction with a hierarchical Bayesian
framework.

Ultimately, the source of the discrepancy between the performance of the two
contrast mechanisms in category generation is still not entirely clear. Our
results could be an indication of a fundamental difference between the processes
involved where exemplar dissimilarity is a larger factor under certain category
configurations while representativeness exerts more influence in others.
Alternatively, it is possible they are both approximating some factor that we
still have not fully identified. While a thorough investigation is beyond the
scope of the current article, the preliminary results of this section indicate
that further work to disentangle these two mechanisms would be promising. In the
following section, we explore some other limitations of the current work and
provide more recommendations for future research.

@@

@@

\subsection{Finding Contrast Effects in Prior Work}

Although existing accounts of category generation broadly overlook the role of
category contrast in determining what is novel versus familiar, it was
implicitly assumed that learners in previous experiments were {\em successful}
in creating new categories. Thus, if contrast plays a robust role in category
genration, its effects should be discernable within the experimental results of
these studies. To provide a test of the influence of category contrast within
existing data, we conducted a novel analysis of Experiment 3 from
\cite{jern2013probabilistic}.\footnote{Although \cite{jern2013probabilistic}
	reported four experiments, we focus on Experiment 3 because their first two
	experiments tested generation of items into known categories, and their fourth
	experiment was identical to Experiment 3 but with a restricted generation
	space.}

Participants in their experiment were exposed to members of two
experimenter-defined categories of `crystals' varying in hue, saturation, and
size. Each category possessed a unique hue, but varied in saturation and size:
In the `Positive' condition, there was a positive correlation between these
features (i.e., larger sized crystals were more saturated), and in the
`Negative' condition this relation was reversed. In the `Neutral' condition,
there was no correlation between saturation and size. After learning about the
categories from each condition, participants were asked to generate six
exemplars belonging to a novel class. As noted above,
\cite{jern2013probabilistic} found that the generated categories tended to
follow the distributional properties of the experimenter-defined categories:
Generated categories were tightly distributed along the hue feature, and
possessed the same saturation-size correlations as in the learned categories.
\cite{jern2013probabilistic}, however, did not analyze or discuss how the
generated categories {\em differed} from the experimenter-defined categories.

Because each experimenter-defined category in the \cite{jern2013probabilistic}
experiment possessed a distinct hue shared by all members of the category, it is
sensible that participants might generate a category with a hue distinct from
the experimenter-provided categories. If category generation were influenced by
category contrast in this way, the hues of generated categories should be
systematically different from those of the experimenter-defined categories.
Unfortunately, stimulus hue was encoded and presented in the
Hue-Saturation-Value (HSV) color space, which is device-dependent and not
perceptually normed such that perceived color similarity corresponds to
proximity in the color space (as opposed to a color space such as CIELAB that is
device-independent and equidistant sets of points correspond to pairs of colors
that have the same perceptual similarity; \citealp{wyszecki1967}). Further, they
did not calibrate their monitor, and so we cannot know the precise colors
presented to participants. As \cite{jern2013probabilistic} were interested in
relations between the saturation and length of examples in generated into novel
categories, these issues do not undercut their analyses and results. However,
these issues pose a significant challenge to evaluating contrast between the
experimenter-defined and participant-generated categories along the hue
dimension. It is plausible that two uncalibrated monitors could display the same
HSV color and the colors be perceived in different color categories (especially
for color boundaries that vary over lightness, such as the yellow-brown
boundary).

Although we cannot know the precise colors that were displayed or perceived, we
can still analyze their results from a coarse perspective to see whether there
is preliminary support for contrast. To do so, we binned all possible hues into
one of eight uniformly-spaced color groups: \{{\em Red}: $0-0.063$, $0.938-1$,
{\em Yellow}: $0.063-0.188$, {\em Yellow-Green}: $0.188-0.313$, {\em
	Green-Teal}: $0.313-0.438$, {\em Teal}: $0.438-0.563$, {\em Teal-Blue}:
$0.563-0.688$, {\em Purple}: $0.688-0.813$, {\em Pink}: $0.813-0.938$\}. In the
\cite{jern2013probabilistic} experiment, the hue of each experimenter-defined
category was selected from one of six possible values, each of which falls into
one of the color groups above (two color groups were not used as a possible hue
for the experimenter-defined categories). By categorizing the
participant-generated crystals likewise, we can obtain a broad measure of
category contrast by determining the proportion of participant-generated
crystals that fall into the same groups as the experimenter-defined categories:
If contrast influences the hues of the generated categories, we should observe
minimal overlap between in the color groupings.

\begin{figure}
	\begin{center} \inputpgf{figs/}{jk13-huecontrast.pgf}
		\caption{Analysis of data from \cite{jern2013probabilistic}, Experiment
			3. Plotted is the number of generated items that share a color group with one of
			the experimenter-defined classes. The ``Expected'' data follows a Binomial
			distribution with $p = 2/8 = 1/4$, given there were two experimenter-defined
			classes, and eight color groups.}
		\label{fig:jk13-huecontrast}
	\end{center}
\end{figure}

These data, shown in Figure \ref{fig:jk13-huecontrast}, reveal a clear pattern:
The majority of participants in each condition ($n = 22$) generated categories
possessing entirely distinct hues; with 0/6 exemplars sharing a hue with the
experimenter-defined categories. These results can be compared to the
predictions of a Binomial model, which proposes that participants generate hues
at random. That is, if hue selection is not systematic, the probability that any
given example will lie in the same color group as an experimenter-defined
category is given by a Binomial distribution with $p = 2/8=1/4$, as there were
two experimenter-defined categories and eight possible color groups. Chi-square
goodness-of-fit tests reveal that the observed distribution in each condition is
highly inconsistent with the hues being chosen at random, all
$\chi^2(6,N=22)>200$, $p < .001$. Participants tended to generate items that were
perceptually distinct from the categories they had learned, and were less likely
to generate hues possessed by members of the experimenter-defined categories.

It is possible, however, that this data can be explained by a process that does
not involve contrast. Specifically, since participants were trained on exemplars
that share the same hue within a category, it is plausible that they generate
exemplars that all share a common hue. If this were the case, given a $6/8 = .75$
probability that a hue distinct from the experimenter-defined categories is
selected, there is a corresponding $.75$ probability (or about 17 out of 22
participants) that the generated category will have no exemplars that share a
hue with any  of the experimenter-defined categories. In addition, that there is
a $2/8 = .25$ probability (about 6 out of 22 participants) that the generated
category will have all six exemplars that share a hue with any of the
experimenter-defined categories. This prediction of regularity across categories
is much closer to the data than the prediction from a Binomial model and does
not require any mechanism of category contrast.

However, upon analyzing the consistency of hues in the generated categories, it
is clear that participants do not generally produce categories with entirely
similar hues across all exemplars. Specifically, only 55\%  of all generated
categories comprised exemplars with a common hue. Consequently, the assumption
of regularity across categories may be inappropriate for this data set, leaving
the contrast-based explanation as the most plausible for our purposes.

Re-analyzing the results from \cite{jern2013probabilistic} provides some
corroborating support for contrast playing a role in category generation. Taken
alongside the analyses reported by \cite{jern2013probabilistic}, our analysis
suggests that generated categories tend to be distinct from {\em and}
distributionally similar to what is already known. However, it is worth noting
that our analysis is still limited: The color groups defined above are
imprecise, and it is not clear that our color grouping is consistent with the
psychological color boundaries perceived by participants. While we did obtain
similar results using a variety of alternative groupings, the hue dimension used
in the \cite{jern2013probabilistic} study does not lend itself straightforwardly
to the computation of similarities, and thus we cannot be certain of whether our
coding accurately approximates the psychological space of the stimuli. This
precludes traditional applications of categorization models to their data as it
is usually necessary to encode objects in psychological space in order to
accurately determine the similarity between objects. By consequence, although
these results likely indicate that contrast exerts {\em some} influence, they do
not precisely describe the nature of that influence.

>>>>

To our knowledge, no prior literature has yet explicitly discussed any empirical
effects of contrast in category generation. However, some evidence of contrast
can be seen in the data from Experiment 3 of cite{jern2013probabilistic}. We
leave a detailed analysis of this data to the supplemental materials.

@@

Despite both the original hierarchical Bayesian model and our representativeness
model sharing identical hierarchical structures, their distinct response
processes can yield very different exemplars. Specifically, the hierarchical
Bayesian model emphasizes the generation of exemplars that maintain
distributional commonalities across categories. In contrast, the
representativeness model focuses on the generation of exemplars that are
representative of the underlying distribution for a target category. Generally,
with the assumption of unimodal underlying distributions, this mechanism of
representativeness results in the generation of novel exemplars that are less
similar to exemplars from the known category. This occurs because the
representativeness of novel exemplars tend to be highest where the probability
density of the known category's underlying distribution is the lowest. We see
this illustrated in Figure \ref{fig:representative-examples}, where the
representativeness model displays a strong preference for generating exemplars
from target category 'B' that are further away from the contrast category 'A'.
Exceptions to this pattern can occur (i.e., the generation of novel exemplars
similar to known category exemplars) -- we explore this in more detail in Section
\ref{section:diffcon} where we highlight an important condition demonstrating
how PACKER and the representativeness model can make qualitatively different
predictions.

@@

For the representativeness model, adding an unbounded feature changes the
probability distribution for the underlying Bayesian model. Whereas it is still
Gaussian in one feature, the other feature is toroidal, meaning that it wraps
back to $0$ radians once it reaches $2\pi$ (we have conducted a change of
variables where $y \rightarrow y \mod 2\pi $, where $\mod$ is the modulus
operator). As we cross $2\pi$ and reach $0$ again, the remaining probability
density also gets wrapped around, meaning that the probability density assigned
by the Gaussian distribution on the toroidal dimension at $2\pi$ is actually
assigned to $0$ (and $2\pi + \epsilon$ to $\epsilon$, and so on). This is called the
{\em wrapped} Gaussian distribution \citep{mardia1972}. Note that the often used
distribution for circular dimensions, the Von Mises distribution, is an
approximation to the wrapped Gaussian distribution. See Appendix B for
implementational details of the representativeness model in toroidal space.

@@

We recruited 89 participants from Amazon Mechanical Turk who were randomly
assigned to the bounded, square stimuli ($N=46$), or to the toroidal, circle
stimuli ($N=43$). The square stimuli were exactly the same as in Experiments 1
and 2. The circle stimuli were made up of unfilled circles with a radial line
extending from its center to its edge. These stimuli varied along two features
-- their diameter ranging from 3.0--5.8 cm, and their orientation ranging from
$2.587$--$2\pi + 2.587$ radians, with $0$ radians indicating a completely
horizontal line. The line widths of the circles and radial line were set to 0.05
cm. Similar to Experiments 1 and 2, participants first observed a training phase
where four Alpha exemplars were presented across three blocks followed by the
generation phase. The training condition was the Corner category for all
participants.

@@

The contrast models were fit by optimizing their fit to all experiments in this
article (Experiments 1 through 3). Fit statistics and parameter values are
presented in Table \ref{table:corner-model-fits}. Although not our main focus in
this section, we also provide the statistics for the classical models --
unsurprisingly, they do not perform as well as the contrast models. Overall,
the representativeness model emerges as the best-fitting model in aggregate, although by a much smaller amount than Experiments 1 and 2. However, analyzing data in aggregate can be misleading.

@@

\begin{tabular}{ l l l l} \\ \\ \textbf{PACKER} & \textbf{Copy \& Tweak} &
\textbf{Representativeness} & \textbf{Hierarchical Bayesian} \\ 
  \hline $AIC = 12023$ &$AIC = 12779$ &$AIC = 11982$  &$AIC = 12976$\\ 
  $L = -6008$          &$L = -6388$   &$L = -5987$    &$L = -6484$\\ 
  $c = 0.51$           &$c = 3.24$    &$\kappa = 6.73$&$\kappa = 0.00$  \\ 
  $\theta_c = 3.03$    &$\theta_c = 0$ (fixed)  &$\nu = 1.00$ &$\nu = 3.27$\\
  $\theta_t = 3.46$    &$\theta_t = 2.90$       &$\lambda = 1.90$ &$\lambda = 0.07$\\
                       &              &$\theta = 2.45$ &$\theta = 3.15$\\
\end{tabular}
\end{table}

@@

We conducted simulations comparing the contrast models' account of our results
to the classical proposals for category generation: a ``copy-and-tweak'' model
(realized as a variant of PACKER with no sensitivity to category contrast), and
a hierarchical Bayesian model designed to explain the classic distributional
similarity effect. In all simulations, we found that the contrast models
captured a previously unexplained and unexplored aspect of human category
generation. In particular, by measuring PACKER's fit as a function of its
prioritization of within- and between-category similarity, we observed that
considering either constraint exclusively results in a relatively low-quality
account. Instead, PACKER's best results were obtained when both constraints are
considered, indicating that human learners do not generate novel concepts
exclusively on the basis of within-category similarity or between
class-contrast. This finding mirrors our behavioral results and demonstrates
that both constraints influence generation when explained with an exemplar
model.

While we found that the representativeness model was found to be a better fit to
data from Experiments 1 and 2, both contrast models essentially made
qualitatively similar predictions across all conditions in those experiments. In
Experiment 3, we collected behavioral data on a category generation condition
that qualitatively distinguished the two contrast models. Specifically, in this
condition where exemplars were located at the corners of the stimulus space,
PACKER predicted the initial generation of centrally-located novel exemplars
while the representativeness model predicted novel exemplars similarly located
at the corners. Interestingly, data from this experiment was better fit,
qualitatively and quantitatively, by PACKER than the representativeness model.

% Ultimately, our category generation data were best explained by the
% representativeness model. The exact nature of its superiority over the other
% models is still not yet clear, although we find evidence that it only barely
% surpasses PACKER when considering known categories with extreme exemplars.

@@

How well does a representativeness model that uses exemplar similarity to
represent the category’s density over features explain our experimental results?
We find that for data across all experiments, this model substantially
outperforms its classical counterpart, but not to the extent of PACKER or the
representativeness model. Specifically, the model fits yielded $L = -4815$ and
$AIC=9633$ for the entire set of data from Experiments 1 and 2, and $L = -3474$,
$AIC=6953$ for the same data excluding the first trials. When we fit the model
to data from Experiments 1,2 and 3, we find that $L = -6301$, $AIC = 12606$.
These results suggest that the predictive advantages to including the
representativeness mechanism is not limited to a hierarchical Bayesian model,
but can also be found when applied to an exemplar model. Although its
performance is not as strong as either of the fully-developed contrast models,
there is a benefit to the representativeness mechanism that is independent from
any interaction with a hierarchical Bayesian framework.

Ultimately, it appears that there is no single model that can capture the
entirety of our empirical data. While the representativeness model, and indeed
the representativeness mechanism itself, offers a predictive advantage in
category generation across all out models, PACKER performs better when
accounting for data from the Corner conditions. While we have established that contrast
mechanisms are essential in modeling category generation, there is no conclusive
evidence that one form of contrast is superior to the other across all
conditions. 
%Xian: Consider rewriting paragraph if there's time.

@@

\newpage
\section{Representativeness Model in Toroidal Space.}
\label{ap:wrapgauss}

Assuming a single bounded feature, we can easily represent the probability
density of a function at any point $x$ on a line as $f(x)$. With a single
unbounded feature, the points $x$ are mapped onto a unit circle with the
corresponding $f(x)$ ``wrapped'' around the circle. More formally, we can
represent the probability density along a wrapped axis $f'(x)$ as:
\begin{align}
f'(x) &= ... + f(x-4\pi) + f(x-2\pi) + f(x) + f(x+2\pi) + f(x+4\pi) + ...\\
     &= \sum_{i=-\infty}^{\infty} f(x + 2\pi i)
\end{align}

We implement the wrapped axis in the representativeness model when
calculating the multivariate Gaussian components of the representativeness equation
$R(x,h)$ (Equation \ref{eq:representativeness} of the main text.) Specifically,
the probability density of the Gaussian component along a single wrapped axis can be computed as:
\begin{align}
p'(x|h) = \sum_{i=-\infty}^{\infty} p(x + 2\pi i|h)
\end{align}
where $p(.|h)$ represents the probability density of the Gaussian component $h$
on a line (i.e., not wrapped around a unit circle.)

In practice, we can approximate the infinite summation by constraining $i$
within a limited range (e.g., $[-50,50]$). For increased efficiency, we can also
approximate $p'(x|h)$ by stopping the summation once the ratio of the maximum
and minimum values of $p(x + 2\pi i|h)$ computed thus far is large enough (e.g.,
10,000.) For our modeling exercises here we impose both contraints, although the
first constraint shows some redundancy since we find that $i$ rarely, if ever,
exceeds a value of 4. Given the unimodal nature of the Gaussian distribution,
this procedure allows for comparatively efficient approximation of the wrapped
distribution without needing to compute the infinite sum.

@@
